Improving Particle Swarm Optimization \\ in a Parallel World

### 2009-04-12 honors_thesis/thesis/paper.tex
A draft of my paper for English 316 is done.

The rest of this paper is outlined as follows.  In Section~\ref{sec:pso}, I
formally describe the particle swarm optimization algorithm, after giving a
brief description of optimization in general.  In Section~\ref{sec:topology},
I describe the sociological aspect of particle swarm optimization and
improvements that we have made for large swarms.  Section~\ref{sec:specex} then
introduces the idea of speculative execution in general and shows how it can be
done in particle swarm optimization.  Section~\ref{sec:results} shows results
of experiments run, and in Section~\ref{sec:conclusion} I draw some
conclusions.

@@

Speculative execution in a program is the execution of code that may or may not
end up actually being needed.  Modern processors routinely do this when a
conditional branch is encountered---they try to predict which branch will
actually be needed and continue their execution.  If it turns out that the
prediction was incorrect, the work is discarded.  But if the prediction was
right, the program can execute much faster than if it had waited on the branch.  
\subsection{Speculative Execution in Particle Swarm Optimization}

In many optimization algorithms speculative execution is impossible, because
the next iteration of the algorithm depends on the evaluation of the function
during the previous iteration.  In simulated annealing, for instance, the
probability of accepting a new position depends on the difference in function
value between the new position and the current one.  In gradient descent, the
gradient of the function must be determined (which includes actually evaluating
it) to know how far to move before sampling the function again.

@@

It uses a lot of extra work to get the second iteration, but for some functions
performing more iterations makes more progress than adding extra particles.

@@

Thank you for reviewing my proposal.  I look forward to exploring speculative
execution in particle swarm optimization.

>>>>

The rest of this paper is outlined as follows.  In Section~\ref{sec:pso}, I
formally describe the particle swarm optimization algorithm, after giving a
brief description of optimization in general.  In Section~\ref{sec:topology},
I describe the sociological aspect of particle swarm optimization and
improvements that we have made for large swarms.  Section~\ref{sec:specex} then
introduces the idea of speculative execution in general and shows how it can be
done in particle swarm optimization.  I then conclude in Section~\ref{sec:conclusion}.

@@

Speculative execution in a program is the execution of code that may or may not
end up actually being needed.  Modern processors routinely do this when a
conditional branch is encountered---they try to predict which branch will
end up being needed and continue their execution.  If it turns out that the
prediction was incorrect, the work is discarded.  But if the prediction was
right, the program can execute much faster than if it had waited on the branch.

\subsection{Speculative Execution in Particle Swarm Optimization}

In many optimization algorithms speculative execution is impossible, because
the next iteration of the algorithm depends on the evaluation of the function
during the previous iteration.  To know which positions to speculatively
evaluate you have to already have evaluated the current position, and then it's
not speculative anymore.  In simulated annealing, for instance, the probability
of accepting a new position depends on the difference in function value between
the new position and the current one.  In gradient descent, the gradient of the
function must be determined (which includes actually evaluating it) to know how
far to move before sampling the function again.  It may be possible to modify
these algorithms to allow for speculative execution, but their standard
executions cannot be done speculatively.

@@

Speculative execution uses a lot of extra work to compute two iterations at the
same time, but for some functions performing more iterations makes more
progress than adding extra particles.  The No Free Lunch theorem again shows
its head here, though, because if you do better with one algorithm on a certain
function or set of functions, that's because you do worse on other functions.
There are, however, a set of functions that do better with more iterations than
with more particles.

Figure~\ref{fig:sphere} shows an example of this: the function Sphere.  The
graph shows the best function value found over a number of function
evaluations.  This is different than previous graphs that have shown function
value against a number of iterations.  At each iteration, a swarm of 50
particles performs 50 function evaluations, where a swarm of 4000 particles
performs 4000 evaluations.  Figure~\ref{fig:sphere} shows that the extra 3950
function evaluations spent on the extra particles in a 4000 particle swarm
really don't do any good.  It would be more worth it to just do more iterations
of a 50 particle swarm than to have a swarm of 4000 particles.

The benefit of adding more particles comes in when you have hundreds or
thousands of processors in a parallel cluster.  You can't use 4000 machines to
run a swarm of 50 particles---3950 of the processors would be idle the whole
time.  Adding more particles to the swarm then improves performance without
adding any extra time (ignoring inter-processor communication cost for the
moment).  But for some functions, a better idea than using the extra 3950
processors to add more particles to the swarm is to use the processors
speculatively so that two iterations can be done at the same time.
Figure~\ref{fig:sphere} shows just one example of a function that would be
better optimized by performing speculative execution than by adding extra
particles.

\begin{figure}
  \centering
  \includegraphics{evals_sphere.eps}
  \caption{Best value found on the function Sphere versus the number of 
  function evaluations with varying swarm sizes.}
  \label{fig:sphere}
\end{figure}

@@

In a world where people are swimming in processing power, many old algorithms
need to be rethought.  Parallel computing architectures allow for new
techniques that were previously impossible.  In the realm of optimization,
particle swarm optimization is uniquely capable of reaping great benefits from
parallel processing.  I have reviewed just a few ways in which PSO can be
improved to harness the increased power of parallel clusters.  As more
researchers tackle the problems of modifying and adapting algorithms to better
utilize the resources that are now available, our ability to solve problems
with computers will increase dramatically.
